Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-04 index.qmd
section:issues,CaseV; process:write

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level. According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability in the perceived trait level of the stimulus.

@@

@@

<!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
On the one hand, the assumption of *equal dispersion between stimuli* suggests that variability in perceived trait levels is consistent across all stimuli. However, Thurstone observed that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce greater judgment discrepancies due to their unique characteristics. In fact, research on text quality may already provide evidence of the violation of this assumption, particularly through the use of misfit statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. Misfit statistics quantify the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. In this context, texts identified as "misfits" display greater judgment discrepancies than other texts. This implies that the discriminal difference associated with these texts shows a broader dispersion, suggesting that their discriminal processes also exhibit greater variation than other texts. This reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations due to the unique or extreme characteristics of the stimuli.

@@

<!-- Misfiting assignments might be assignments with larger dispersion, because there is a lot of confusion about them. When assessing creativity, for instance, a student might complete an assignment in a highly creative but unexpected way. Judges, unprepared for such originality, may experience uncertainty, leading to greater variability in their judgments about the assignment. Then the discriminal difference is all over the place. -->

@@

<!-- 5. But what are the challenges when these assumptions are violated? -->
[what are the challenges when the equal dispersion assumption is violated?]{style="color:red;"}

<!-- But the statistical literature warns against mis-specifying the trait’s measurement model. When mis-specifying the measurement model by assuming, for example, equal discriminal dispersions across stimuli, the model can overlook meaningful differences in the reliability of the trait. Taken together, these mis-specifications and oversights can lead to misleading conclusions about the trait measurement, ultimately casting doubts about the validity of the trait, as validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. -->

@@

<!-- If your scores are not reliable they cannot be valid. People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid. -->

@@

@@

<!-- This has been considered in the literature as a (see https://doi.org/10.1177/1471082X15571817) or a multilevel BTL model (find sources) -->

@@

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. If your scores are not reliable they cannot be valid. People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. And when you do that, the scores you produce most likely will not be reliable, which, in turn, renders them invalid. -->

@@

>>>>

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level. According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the perceived trait variability of the stimulus.

@@

<!-- Model misspecification happens when the set of probability distributions considered by the statistician does not include the distribution that generated the observed data. -->

@@

<!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
On the one hand, the assumption of *equal dispersion between stimuli* suggests that the perceived trait variability is consistent across all stimuli. However, Thurstone observed that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics. Research on text quality may already provide evidence of this assumption's violation, particularly when considering the role of misfit statistics in the analysis of CJ data [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. Misfit statistics quantify the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For instance, texts identified as "misfits" usually exhibit more judgment discrepancies than others. This, in turn, implies that the discriminal difference associated with these texts shows a broader dispersion, ultimately suggesting that their discriminal processes also exhibit more variation than other texts. Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations due to the unique characteristics of the stimuli or the judges themselves.

@@

@@

<!-- 5. But what are the challenges when these assumptions are violated? -->
Moreover, assuming equal dispersion between stimuli despite its violation causes Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli. Moreover, if researchers recognize that misfit statistics help to identify such violations, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020]. The direction and magnitude of these biases remain unpredictable because they depend on which stimuli researchers exclude from the analysis. Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2].

@@

<!-- People have been employing a model with assumptions that they do not know if those assumptions apply for their cases. -->

@@

<!-- Although no CJ literature offer solutions to mitigate the aforementioned issues, we can take lesson from the Structural Equation Model (SEM) literature. More specifically, from @Depaoli_2021. -->

@@

@@

<!-- Reliability is a necessary but not sufficient condition for validity. Reliability can exist without validity but validity cannot exist without reliability [@Perron_et_al_2015, pp. 2]. -->

@@

<!-- include subject specific covariates [@Casalicchio_et_al_2015] -->
